\systemEnding: Bootstrapping Domain-Specific Languages for Neurally-Guided Bayesian Program Learning

### 2019-04-17 archive.tex
improved week/sleep diagram
more progress on physical laws

\begin{figure}
  \begin{tikzpicture}[line width=0.4mm]
    \draw[fill=teal!5!white] (-1,1) -- (13,1) -- (13,-4) -- (-1,-4) -- (-1,1);
    \node at (5.5,1.25) {\textsc{\textbf{Wake}}};

    \begin{scope}[shift={(0.5,0.5)}]
      \node[align=center] at (0,-0.5) (d){\textbf{Library}\\
        \code{$f_1($x$)=$(+ x 1)}\\
        \code{$f_2($z$)=$(fold cons (cons z nil))}\\
         ...};
      \node[align=center] at ([yshift = -2cm]d) (t){\textbf{Task}\\
        \footnotesize            \code{[7\, 2\, 3]}$\to$\code{[4 3 8]}         \\
        \footnotesize    \code{[3\, 8]}$\to$\code{[9 4]}\\
        \footnotesize    \code{[4\, 3\, 2]}$\to$\code{[3 4 5]} };

@@

      \draw [->  ,rounded corners] (t.south) -- ([yshift = -0.5cm]t.south) -- ([yshift = -0.5cm] s.south |- t.south) -- (s.south);
    \end{scope}
    \begin{scope}[shift={(9.4,-3.5)},scale=0.6,line width=0.05mm]
      \node[obs,scale=0.7] at (3.5,3) (dx){Library};
      \node[latent,scale=0.7] at ([yshift=-1.7cm,xshift=0cm]dx) (zp){prog};
      \node[obs,scale=0.7] at ([yshift=-1.45cm]zp) (xp) {task};
      \node[latent,scale=0.7] at ([xshift=1.5cm]zp) (zp1){prog};
      \node[obs,scale=0.7] at ([xshift=1.5cm]xp) (xp1) {task};
      \draw [->] (zp1.south) -- (xp1.north);
      \draw [->] (dx.south) -- (zp1.north);
      \draw [->,red] (xp1.east) to[out = 30,in = -30] node(nn){} (zp1.east);
      \node[latent,scale=0.7] at ([xshift=-1.5cm]zp) (zp1){prog};
      \node[obs,scale=0.7] at ([xshift=-1.5cm]xp) (xp1) {task};
      \draw [->] (zp1.south) -- (xp1.north);
      \draw [->] (dx.south) -- (zp1.north);
      \draw [->] (dx.south) -- (zp.north);
      \draw [->] (zp.south) -- (xp.north);
      \draw [->,red] (xp1.east) to[out = 30,in = -30] node(nn){} (zp1.east);
      \draw [->,red] (xp.east) to[out = 30,in = -30] node(nn){} (zp.east);
    \end{scope}

@@

Rather than start with a domain-specific basis and then enrich it over successive wake/sleep cycles -- as these prior experiments have done --
could we instead start with a highly generic basis,
and then learn a library encoding something closer to the basic domain language?
We first consider learning a basic language for equations describing physical laws,
starting \emph{without} concepts like vectors or forces
and instead providing Lisp-style sequence manipulation primitives like
\code{map} and \code{fold}.
We collected 57 equations for physical laws from AP and MCAT physics `cheat sheets' --
forms like Coulomb's law and expressions for the center of mass of a set of objects --
along with
reference vector algebra identities from machine learning textbooks,
and tasked our agent with inferring programs calculating these input/output mappings.
In solving these tasks, \system first discovers the building blocks of vector algebra,
like inner products, vector sums, and norms,
and then uses this vocabulary to infer library routines for common
physical relationships like inverse square laws and integrating second derivatives over time.

\begin{figure}
  \centering
  \begin{tabular}{lll}\toprule
    &\multicolumn{1}{c}{Physical Laws}&\multicolumn{1}{c}{Recursive Functions}\\\midrule
    \normalsize \rotatebox[origin=c]{90}{\pop{Programs} \& Tasks}&
    % physical laws programs and tasks
    \begin{tabular}{l}
      \begin{tabular}{l}
        Center-of-mass:\\
      \end{tabular}\\\\
      \begin{tabular}{l}
        Coulomb's law:\\
      \end{tabular}\\\\
      \begin{tabular}{l}
        Integrating position over time:\\
        \end{tabular}
      \end{tabular}
    &
    % recursive functions programs and tasks
    \begin{tabular}{l}
      \begin{tabular}{l}
        Duplicate each list element:\\
        \code{[9 2]$\to $[9 9 2 2]}\\
        \code{[5 0 3]$\to $[5 5 0 0 3 3]}\\
        \code{[1 2 3 4]$\to $[1 1 2 2 3 3 4 4]}\\
        \blueCode{$f(l) = $($f_0$ ($\lambda$ (x a) (cons x (cons x a))) l nil)}        
      \end{tabular}\\\\

@@

      \end{tabular}
    \\\midrule
    \rotatebox[origin=c]{90}{\popp{Library}}&
    % physical laws library
    &
    % recursive functions programs and tasks
    \begin{tabular}{l}
      \begin{tabular}{l}
        \popp{\code{$f_0($f$,$l$,$x$)\,=\,$(if (empty? l) x}}\\
        \phantom{\code{$f_0($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(f (car l) ($f_0$ (cdr l))))}}\\
          \hspace{\helpSize}($f_0$: \emph{fold})
      \end{tabular}\\
      \begin{tabular}{l}
        \popp{\code{$f_1($p$,$f$,$n$,$x$)\,=\,$(if (p x) nil}}\\
        \phantom{\code{$f_1($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(cons (f x) ($f_1$ (n x))))}}\\
          \hspace{\helpSize}($f_1$: \emph{unfold})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_2($f$,$l$)\,=\,$($f_0$ nil l ($\lambda$ (x a) (cons (f x) a)))}\\
        \hspace{\helpSize}($f_2$: \emph{map})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_3($f$,$p$,$n$)\,=\,$($f_1$ p f (+ 1) n)}\\
        \hspace{\helpSize}($f_3$: \emph{count upward until predicate holds})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_4($l$)\,=\,$($f_0$ ($\lambda$ (a x) (+ 1 a)) l 0)}\\
        \hspace{\helpSize}($f_4$: \emph{length})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_5($p$)\,=\,$($f_3$ ($\lambda$ (x) x) p 0)}\\
        \hspace{\helpSize}($f_5$: \emph{count upward from 0 until predicate holds})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_6($n$)\,=\,$($f_5$ (eq? n))}\\
        \hspace{\helpSize}($f_6$: \emph{range})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_7($l$,$n$)\,=\,$(car ($f_0$ ($\lambda$ (a x) (cdr a)) ($f_6$ n) l))}\\
        \hspace{\helpSize}($f_7$: \emph{index})
      \end{tabular}

@@

  \end{tabular}
  \end{figure}

Could we also learn the basics of programming, including these foundational sequence manipulation primitives?
This goal is relevant to a long-standing dream within the program induction community to ``learn from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal,schmidhuber2004optimal,hutter2004universal,solomonoff1989system}.
All existing systems,
including ours,
fall far short of this dream,
and it is unclear (and we believe unlikely)
that this dream could ever be fully realized.
How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

We ran the following experiment: \system was given a subset of the
1959 Lisp primitives, and tasked with solving 18 programming
exercises. A key difference between this setup and our previous
experiments is that, for this experiment, the system is given
primitive recursion, whereas previously we had sequestered recursion
within higher-order functions like \code{map}, \code{fold}, and
\code{unfold}.

After running for 93 hours on 48 CPUs, our
algorithm solves these 18 exercises, along the way assembling a DSL
with a modern repertoire of
functional programming idioms and subroutines, including \code{map},
\code{fold}, \code{unfold}, \code{index}, \code{length},
and  arithmetic operations like 
building lists of natural numbers between an interval (see  Appendix~\ref{appendixMcCarthy}).

We believe that program learners should \emph{not}
start from scratch,
but instead should start from
a rich, domain-agnostic
basis like those embodied in the standard libraries of modern  languages.
What this experiment shows is that \system doesn't \emph{need} to start from a rich basis,
and can in principle recover many of the amenities of modern programming systems,
provided it is given enough computational power and a suitable
spectrum of tasks.

@@

\subsection{Learning from Scratch: Tasks and DSL}\label{appendixMcCarthy}

@@

Figure~\ref{learningFromScratch} shows the full set of tasks and the learned DSL.
\begin{figure}  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{ll}\toprule
    \normalsize \rotatebox[origin=c]{90}{\pop{Programs} \& Tasks}&%&\normalsize \popp{DSL}\\\midrule

@@

      \begin{tabular}{l}
        \code{0, [9 2 3]$\to $9}\\
        \code{3, [0 2 8 4 5 6]$\to $4}\\
        \blueCode{$f($n$,$l$) = $($f_{10}$ l n)}
      \end{tabular}&
      \begin{tabular}{l}
        \code{1, [9 2 3]$\to $9}\\
        \code{4, [0 2 8 1 5 6]$\to $1}\\
        \blueCode{$f($n$,$l$) = $($f_{10}$ l (+ 1 n))}
      \end{tabular}\\\\
      \begin{tabular}{l}
        \code{3$\to $[-3 -2 -1]}\\
        \code{4$\to $[-4 -3 -2 -1]}\\
        \blueCode{$f($n$) = $($f_{8}$ 0 n)}
      \end{tabular}&
      \begin{tabular}{l}
        \code{2$\to $[2 1]}\\
        \code{4$\to $[4 3 2 1]}\\
        \blueCode{$f($n$) = $($f_{8}$ 0 (- 0 n))}
      \end{tabular}      
    \end{tabular}
    \\\midrule 
    \rotatebox[origin=c]{90}{\normalsize \popp{DSL}}&
    \begin{tabular}{ll}
      \begin{tabular}{l}
        \popp{\code{$f_0($f$,$l$,$x$)\,=\,$(if (empty? l) x}}\\
        \phantom{\code{$f_0($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(f (car l) ($f_0$ (cdr l))))}}\\
          \hspace{\helpSize}($f_0$: \emph{fold})
      \end{tabular}&
      \begin{tabular}{l}
        \popp{\code{$f_1($p$,$f$,$n$,$x$)\,=\,$(if (p x) nil}}\\
        \phantom{\code{$f_1($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(cons (f x) ($f_1$ (n x))))}}\\
          \hspace{\helpSize}($f_1$: \emph{unfold})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_2($f$,$l$)\,=\,$($f_0$ nil l ($\lambda$ (x a) (cons (f x) a)))}\\
        \hspace{\helpSize}($f_2$: \emph{map})
      \end{tabular}&
      \begin{tabular}{l}
                  \greenCode{$f_3($f$,$l$)\,=\,$($f_0$ nil l ($\lambda$ (x a) (if (f x) a (cons x a))))}\\
          \hspace{\helpSize}($f_3$: \emph{filter})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_4($f$,$p$,$n$)\,=\,$($f_1$ p f (+ 1) n)}\\
        \hspace{\helpSize}($f_4$: \emph{count upward until predicate holds})
      \end{tabular}&
      \begin{tabular}{l}
        \greenCode{$f_5($l$)\,=\,$($f_0$ ($\lambda$ (a x) (+ 1 a)) l 0)}\\
        \hspace{\helpSize}($f_5$: \emph{length})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_6($n$,$p$,$l$)\,=\,$($f_1$ p n car l)}\\
        \hspace{\helpSize}($f_6$: \emph{specialization of unfold})
      \end{tabular}&
      \begin{tabular}{l}
        \greenCode{$f_7($p$)\,=\,$($f_4$ ($\lambda$ (x) x) p 0)}\\
        \hspace{\helpSize}($f_7$: \emph{count upward from 0 until predicate holds})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_8($n$,$m$)\,=\,$($f_4$ ($\lambda$ (x) (- n x)) (eq? 0) m)}\\
        \hspace{\helpSize}($f_7$: \emph{count downwards})
      \end{tabular}&
      \begin{tabular}{l}
        \greenCode{$f_9($n$)\,=\,$($f_7$ (eq? n))}\\
        \hspace{\helpSize}($f_9$: \emph{range})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_{10}($l$,$n$)\,=\,$(car ($f_0$ ($\lambda$ (a x) (cdr a)) ($f_9$ n) l))}\\
        \hspace{\helpSize}($f_{10}$: \emph{index})
      \end{tabular}&

>>>>

\begin{figure}
  \begin{tikzpicture}[line width=0.4mm]
    \draw[fill=teal!5!white] (-1.25,1.25) -- (13.25,1.25) -- (13.25,-4) -- (-1.25,-4) -- (-1.25,1.25);
    \node at (5.5,1.5) {\textsc{\textbf{Wake}}};

    \begin{scope}[shift={(0.5,0.5)}]
      \node[align=center] at (0,-0.5) (d){
        \begin{tabular}{l}
          \multicolumn{1}{c}{\textbf{Library}}\\
        \small\code{$f_1($x$)=$(+ x 1)}\\
        \small\code{$f_2($z$)=$(fold cons}\\
        \small\phantom{\code{$f_2($z$)$}}\code{(cons z nil))}\\
\small $\cdots\cdots\cdots$
                \end{tabular}};
      \node[align=center] at ([yshift = -2cm]d) (t){\textbf{Task}\\
        \footnotesize            \code{[7\, 2\, 3]}$\to$\code{[4 3 8]}         \\
        \footnotesize    \code{[3\, 8]}$\to$\code{[9 4]}\\
        \footnotesize    \code{[4\, 3\, 2]}$\to$\code{[3 4 5]} };

@@

      \draw [->  ,rounded corners] (t.south) -- ([yshift = -0.5cm]t.south) -- ([yshift = -0.5cm] s.south |- t.south) -- (s.south);
    \end{scope}

    \begin{scope}[shift={(9.4,-3.5)},scale=0.6,line width=0.05mm]
      \node[obs,scale=0.7] at (3.5,3) (dx){Library};
      \node[latent,scale=0.7] at ([yshift=-1.7cm,xshift=0cm]dx) (zp){prog};
      \node[obs,scale=0.7] at ([yshift=-1.45cm]zp) (xp) {task};
      \node[latent,scale=0.7] at ([xshift=1.5cm]zp) (zp1){prog};
      \node[obs,scale=0.7] at ([xshift=1.5cm]xp) (xp1) {task};
      \draw [->] (zp1.south) -- (xp1.north);
      \draw [->] (dx.south) -- (zp1.north);
      \draw [->,red] (xp1.east) to[out = 30,in = -30] node(nn){} (zp1.east);
      \node[latent,scale=0.7] at ([xshift=-1.5cm]zp) (zp1){prog};
      \node[obs,scale=0.7] at ([xshift=-1.5cm]xp) (xp1) {task};
      \draw [->] (zp1.south) -- (xp1.north);
      \draw [->] (dx.south) -- (zp1.north);
      \draw [->] (dx.south) -- (zp.north);
      \draw [->] (zp.south) -- (xp.north);
      \draw [->,red] (xp1.east) to[out = 30,in = -30] node(nn){} (zp1.east);
      \draw [->,red] (xp.east) to[out = 30,in = -30] node(nn){} (zp.east);
    \end{scope}

@@

Rather than start with a domain-specific basis and then enrich it over successive wake/sleep cycles -- as these prior experiments have done --
could we instead start with a highly generic basis,
and then learn a library encoding something closer to the basic domain language?
We first consider learning a basic language for physical laws,
starting with generic sequence-manipulation primitives,
and then investigate how these fundamental sequence-manipulating routines could themselves be learned.

We first task \system with inferring 57 equations describing physical laws,
like closed forms for ballistic motion, inverse square laws,
expressions for calculating centers of mass, etc.,
drawing verbatim from AP and MCAT physics `cheat sheets'
along with equation guides from machine learning textbooks.
These relationships are easy to express
given concepts like vectors or forces.
Rather than `bake in' abstractions like these we
initialize the system with Lisp-style sequence manipulation primitives like
\code{map} and \code{fold}.
We automatically convert each task into a
regression problem by randomly generating real or vector valued input/outputs,
and encode vectors as lists of reals.

In solving these tasks, \system first discovers the building blocks of vector algebra,
like inner products, vector sums, and norms.
It then uses this vocabulary to infer library routines for common
physical relationships like the vector form inverse square laws
and integrating second derivatives over time,
effectively undergoing a `change of basis' from the initial Lisp-style primitives into a
physics-style basis. Figure~\ref{librariesToLanguages}~(left) illustrates a subset
of the provided tasks and learned library for this domain.

\begin{figure}
  \centering
  \begin{tabular}{lll}\toprule
    &\multicolumn{1}{c}{Physical Laws}&\multicolumn{1}{c}{Recursive Functions}\\\midrule
    \normalsize \rotatebox[origin=c]{90}{\pop{Programs} \& Tasks}&
    % physical laws programs and tasks
    \begin{tabular}{l}
      \begin{tabular}{l}
        Center-of-mass:\\
        $m_1$, $\vec{r}_1$, $m_2$, $\vec{r}_2\mapsto \frac{m_1}{m_1 + m_2}\vec{r}_1 + \frac{m_2}{m_1 + m_2}\vec{r}_2$\\
        \code{2, [3 0 1], 4, [0 6 0]}$\to$\code{[1 4 0.333]}\\
        \code{5, [7 1 6], 5, [9 2 2]}$\to$\code{[8 0.5 4]}        
      \end{tabular}\\\\
      \begin{tabular}{l}
        Coulomb's law:\\
        $q_1$, $\vec{r}_1$, $q_2$, $\vec{r}_2\mapsto \frac{q_1q_2}{|\vec{r}_1 - \vec{r}_2|_2^2} \times \frac{\vec{r}_1 - \vec{r}_2}{|\vec{r}_1 - \vec{r}_2|}$\\
        \code{4, [1 5 2], 2, [2 5 2]$\to$[-8 0 0]}\\
        \code{-1, [0 1 1], -1, [0 0 0]$\to$[0 0.5 0.5]}        
      \end{tabular}\\\\
      \begin{tabular}{l}
        Integrating position over time:\\
        $x_0$, $\dot{x}$, $\ddot{x}$, $t\mapsto \frac{1}{2}\ddot{x}t^2 + \dot{x}t + x_0$\\

        \end{tabular}
      \end{tabular}
    &
    % recursive functions programs and tasks
    \begin{tabular}{l}
      \begin{tabular}{l}
        Duplicate each list element:\\
        \code{[9 2]$\to $[9 9 2 2]}\\
        \code{[5 0 3]$\to $[5 5 0 0 3 3]}\\
        \code{[1 2 3 4]$\to $[1 1 2 2 3 3 4 4]}\\
        \blueCode{$f(l) = $($f_0$ ($\lambda$ (x a) (cons x (cons x a))) l nil)}        
      \end{tabular}\\\\

@@

      \end{tabular}
    \\\midrule
    \rotatebox[origin=c]{90}{\popp{Library}}&
    % physical laws library
    \begin{tabular}{l}
      \begin{tabular}{l}
        \greenCode{$f_0($v$)\,=\,$(fold 0 ($\lambda$ (x y) (+ (* x x) y)) v)}\\
        \hspace{\helpSize}($f_0$: \emph{L-2 norm})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_1($v$)\,=\,$(fold 0 ($\lambda$ (x y) (+ x y)) v)}\\
        \hspace{\helpSize}($f_1$: \emph{Sum vector components})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_2($v,u$)\,=\,$($f_1$ (zip v u ($\lambda$ (x y) (* x y))))}\\
        \hspace{\helpSize}($f_2$: \emph{Dot product})
      \end{tabular}\\
      %(lambda (fold $0 (map (lambda (*. $0 0.)) (car $0)) (lambda (lambda (zip $0 $1 (lambda (lambda (+. $0 $1)))))))) 
      \begin{tabular}{l}
        \greenCode{$f_3($vs$)\,=\,$(fold vs (map ($\lambda$ (x) (* x 0))}\\
        \phantom{\greenCode{$f_3($vs$)\,=\,$(fold vs (}}\greenCode{(car vs))}\\
        \phantom{\greenCode{$f_3($vs$)\,=\,$(}}\greenCode{($\lambda$ (v u) (zip v u ($\lambda$ (x y) (+ x y)))))}\\
        \hspace{\helpSize}($f_3$: \emph{Sum (list of) vectors})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_4($x$)\,=\,$(/ x (+ 1 1))}\\
        \hspace{\helpSize}($f_4$: \emph{Divide by two})
      \end{tabular}\\
      %      %#(lambda (lambda (lambda (lambda (+. $2 (*. $0 (+. (*. (#(lambda (/. $0 (+. 1. 1.))) $0) $1) $3))))))) $2 $3 $1 $0)))))
      \begin{tabular}{l}
        \greenCode{$f_5($v,x,a,t$)\,=\,$(+ x (* t (+ (* ($f_4$ t) a) v)))}\\
        \hspace{\helpSize}($f_5$: \emph{Integrate given first and second derivatives})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_6($v$)\,=\,$(power ($f_0$ v) ($f_4$ 1))}\\
        \hspace{\helpSize}($f_6$: \emph{n-dimensional vector length})
        \end{tabular}
      \end{tabular}
    &
    % recursive functions programs and tasks
    \begin{tabular}{l}
      \begin{tabular}{l}
        \popp{\code{$f_0($f$,$l$,$x$)\,=\,$(if (empty? l) x}}\\
        \phantom{\code{$f_0($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(f (car l) ($f_0$ (cdr l))))}}\\
          \hspace{\helpSize}($f_0$: \emph{fold})
      \end{tabular}\\
      \begin{tabular}{l}
        \popp{\code{$f_1($p$,$f$,$n$,$x$)\,=\,$(if (p x) nil}}\\
        \phantom{\code{$f_1($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(cons (f x) ($f_1$ (n x))))}}\\
          \hspace{\helpSize}($f_1$: \emph{unfold})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_2($f$,$l$)\,=\,$($f_0$ nil l ($\lambda$ (x a) (cons (f x) a)))}\\
        \hspace{\helpSize}($f_2$: \emph{map})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_3($f$,$p$,$n$)\,=\,$($f_1$ p f (+ 1) n)}\\
        \hspace{\helpSize}($f_3$: \emph{count upward until predicate holds})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_4($l$)\,=\,$($f_0$ ($\lambda$ (a x) (+ 1 a)) l 0)}\\
        \hspace{\helpSize}($f_4$: \emph{length})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_5($p$)\,=\,$($f_3$ ($\lambda$ (x) x) p 0)}\\
        \hspace{\helpSize}($f_5$: \emph{count upward from 0 until predicate holds})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_6($n$)\,=\,$($f_5$ (eq? n))}\\
        \hspace{\helpSize}($f_6$: \emph{range})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_7($l$,$n$)\,=\,$(car ($f_0$ ($\lambda$ (a x) (cdr a)) ($f_6$ n) l))}\\
        \hspace{\helpSize}($f_7$: \emph{index})
      \end{tabular}

@@

  \end{tabular}
  \caption{\textbf{Left:} Learning a language for physical laws starting with recursive list routines like \code{map} and \code{fold}. Agent observes input/outputs of latent functions encoding physical laws (top left, 20 input/outputs per task). Vectors are represented as lists of real numbers. Physical constants expressed in Planck units. \textbf{Right:} Learning a language for recursive list routines starting with only recursion and primitives found in 1959 Lisp. Agent rediscovers \code{map}, \code{fold}, \code{unfold}, etc. For full tasks and learned libraries see Appendix~\ref{appendixMcCarthy}}\label{librariesToLanguages}
  \end{figure}

Could we also learn the basics of programming, including these recursive sequence manipulation primitives?
This goal is relevant to a long-standing dream within the program induction community to ``learn from scratch'': starting with a \emph{minimal} Turing-complete programming language,
and then learning to solve a wide swath of
induction problems~\cite{solomonoff1964formal,schmidhuber2004optimal,hutter2004universal,solomonoff1989system}.
All existing systems,
including ours,
fall far short of this dream,
and it is unclear (and we believe unlikely)
that this dream could ever be fully realized.
How far can we push in this direction?
``Learning from scratch'' is subjective, but a reasonable
starting point is the set of primitives provided in 1959
Lisp~\cite{mccarthy1960recursive}: these include
conditionals, recursion, arithmetic, and the 
list operators \code{cons}, \code{car}, \code{cdr}, and \code{nil}.
A  basic first goal is to start with
these primitives,
and then recover a DSL that
more closely resembles modern functional languages like Haskell and OCaml.
Recall (Sec.~\ref{sequences})
that we initially provided our system with functional programming routines like
\code{map} and \code{fold}.

We ran the following experiment: \system was given a subset of the
1959 Lisp primitives, and tasked with solving 18 programming
exercises. A key difference between this setup and our previous
experiments is that, for this experiment, the system is given
primitive recursion, whereas previously we had sequestered recursion
within higher-order functions like \code{map}, \code{fold}, and
\code{unfold}.  In solving these 18 exercises, the algorithm assembles
a DSL containing a modern repertoire of functional programming idioms
and subroutines, including \code{map}, \code{fold}, \code{unfold},
\code{index}, \code{length}, and arithmetic operations like building
lists of natural numbers between an interval (see
Figure~\ref{librariesToLanguages}, right).

We believe that program learners should \emph{not}
start from scratch,
but instead should start from
a rich, domain-agnostic
basis like those embodied in the standard libraries of modern  languages.
What these experiments reveal is that \system doesn't \emph{need} to start from a rich basis,
and can in principle recover many of the amenities of modern programming systems,
provided it is given enough computational power and a suitable
spectrum of tasks.

@@

\subsection{Learning from Scratch: Tasks and Library}\label{appendixMcCarthy}

@@

Figure~\ref{learningFromScratch} shows the full set of tasks and the learned library.
\begin{figure}  \newcommand{\helpSize}{0.25cm}
  \begin{tabular}{ll}\toprule
    \normalsize \rotatebox[origin=c]{90}{\pop{Programs} \& Tasks}&%&\normalsize \popp{DSL}\\\midrule

@@

      \begin{tabular}{l}
        \code{0, [9 2 3]$\to $9}\\
        \code{3, [0 2 8 4 5 6]$\to $4}\\
        \blueCode{$f($n$,$l$) = $($f_{10}$ l n)}
      \end{tabular}&
      \begin{tabular}{l}
        \code{1, [9 2 3]$\to $9}\\
        \code{4, [0 2 8 1 5 6]$\to $1}\\
        \blueCode{$f($n$,$l$) = $($f_{10}$ l (+ 1 n))}
      \end{tabular}\\\\
      \begin{tabular}{l}
        \code{3$\to $[-3 -2 -1]}\\
        \code{4$\to $[-4 -3 -2 -1]}\\
        \blueCode{$f($n$) = $($f_{8}$ 0 n)}
      \end{tabular}&
      \begin{tabular}{l}
        \code{2$\to $[2 1]}\\
        \code{4$\to $[4 3 2 1]}\\
        \blueCode{$f($n$) = $($f_{8}$ 0 (- 0 n))}
      \end{tabular}      
    \end{tabular}
    \\\midrule 
    \rotatebox[origin=c]{90}{\normalsize \popp{Library}}&
    \begin{tabular}{ll}
      \begin{tabular}{l}
        \popp{\code{$f_0($f$,$l$,$x$)\,=\,$(if (empty? l) x}}\\
        \phantom{\code{$f_0($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(f (car l) ($f_0$ (cdr l))))}}\\
          \hspace{\helpSize}($f_0$: \emph{fold})
      \end{tabular}&
      \begin{tabular}{l}
        \popp{\code{$f_1($p$,$f$,$n$,$x$)\,=\,$(if (p x) nil}}\\
        \phantom{\code{$f_1($f$,$l$,$x$)\,=\,$(if }}}\popp{\code{(cons (f x) ($f_1$ (n x))))}}\\
          \hspace{\helpSize}($f_1$: \emph{unfold})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_2($f$,$l$)\,=\,$($f_0$ nil l ($\lambda$ (x a) (cons (f x) a)))}\\
        \hspace{\helpSize}($f_2$: \emph{map})
      \end{tabular}&
      \begin{tabular}{l}
                  \greenCode{$f_3($f$,$l$)\,=\,$($f_0$ nil l ($\lambda$ (x a) (if (f x) a (cons x a))))}\\
          \hspace{\helpSize}($f_3$: \emph{filter})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_4($f$,$p$,$n$)\,=\,$($f_1$ p f (+ 1) n)}\\
        \hspace{\helpSize}($f_4$: \emph{count upward until predicate holds})
      \end{tabular}&
      \begin{tabular}{l}
        \greenCode{$f_5($l$)\,=\,$($f_0$ ($\lambda$ (a x) (+ 1 a)) l 0)}\\
        \hspace{\helpSize}($f_5$: \emph{length})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_6($n$,$p$,$l$)\,=\,$($f_1$ p n car l)}\\
        \hspace{\helpSize}($f_6$: \emph{specialization of unfold})
      \end{tabular}&
      \begin{tabular}{l}
        \greenCode{$f_7($p$)\,=\,$($f_4$ ($\lambda$ (x) x) p 0)}\\
        \hspace{\helpSize}($f_7$: \emph{count upward from 0 until predicate holds})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_8($n$,$m$)\,=\,$($f_4$ ($\lambda$ (x) (- n x)) (eq? 0) m)}\\
        \hspace{\helpSize}($f_7$: \emph{count downwards})
      \end{tabular}&
      \begin{tabular}{l}
        \greenCode{$f_9($n$)\,=\,$($f_7$ (eq? n))}\\
        \hspace{\helpSize}($f_9$: \emph{range})
      \end{tabular}\\
      \begin{tabular}{l}
        \greenCode{$f_{10}($l$,$n$)\,=\,$(car ($f_0$ ($\lambda$ (a x) (cdr a)) ($f_9$ n) l))}\\
        \hspace{\helpSize}($f_{10}$: \emph{index})
      \end{tabular}&
